Efficiently stable presentations from error-correcting codes

### 2023-07-09 efficient_stability.tex
started some comments on Section 5

\begin{lemma}\label{lem:pull-back}
Let  $(\mM,\tau^\mM)$ be a tracial von Neumann algebra, $P\in\mM_\infty$ a projection of finite trace, $\mN=P\mM_\infty P$ and $\tau^\mN=\tau_\infty/\tau_\infty(P)$, and $w\in P \mM_\infty \Id_\mM$ a partial isometry. \hnote{The $P$ here is unrelated to the projective measurement $\{P_a\}$, right? OK if $P$ is changed to, say, $\Pi$? (or alternatively $\{P_a\}$ is named something else.}\tnote{Yes, unrelated. I prefer to keep $P$ because it's used throughout the paper; so I changed $P_a$} Let 
\[ \eps = \max\big\{ \tau^\mM\big(\Id_\mM - w^* w\big)\,,\;\tau^\mN\big( P- w w^*\big)\big\}\;.\] 
 Then for any projective measurement $\{T_a\}_{a \in \mA}$ on $\mN$, there is a projective measurement $\{Q_a\}_{a \in \mA}$ on $\mM$ such that 
\begin{equation}
\label{eq:pull-back} \sum_{a \in \mA} \big\| Q_a - w^* T_a w\big\|_\tau^2 \,\leq \ 56\eps\;.
\end{equation}
\end{lemma}

@@

\begin{definition}[Almost homomorphism]
Let $G = \langle S:R\rangle $ be a finitely presented group, $\mu$ a distribution on $R$, and $(\mM,\tau)$ a tracial von Neumann algebra. An $(\eps,\mu)$-almost homomorphism of $G$ on $(\mM,\tau)$ is a homomorphism $\phi:\mF(S)\to\mU(\mM)$ \hnote{I find it a little funny to call $\phi$ a ``homomorphism'' because the free group doesn't have any relations (maybe except for the identity}\tnote{No but what this means is that e.g.\ $\phi(ab)=\phi(a)\phi(b)$. Equivalently, we could ask $\phi$ to be defined on $S$ only, and then extend it to $\mF(S)$ ``in the natural way''. I preferred to define it on the entire group right away. But there are constaints!} such that
\[ \Es{r\sim \mu} \big\|  \phi(r) - \Id \big \|_\tau^2 \,\leq\, \eps\;.\]
\end{definition}

@@

Let $G_\bRM = G(h_\bRM)$ be the group that is presented from $h_\bRM$. Explicitly, $h_\bRM$ is the parity check matrix that arises from the tester described in Figure~\ref{fig:bRM-tester}.

@@

We do not know if $G_\bRM = \Z_2^K$, with $K=t(d+1)^m$. Instead we modify the presentation $G(h_\bRM)$ by adding pairwise commutation relations in a similar manner as Lemma~\ref{lem:com-code}. Let 
\[ G(h_\bRM) \,=\, \big\langle x_1,\ldots,x_N \;:\; \{R^\sq_k\}\,,\; \{R^\ld_k\}\,,\; \{R^\had_{k}\} \big\rangle\;,\]
where  $R^\sq_k$ ranges overal all relations of the form $x_i^2=e$ for $i\in \{1,\ldots,N\}$, $R^\ld_k$ ranges over all relations implied by the ``low-degree test'' in Figure~\ref{fig:bRM-tester} and $R^\had_{k}$ ranges over all the relations implied by the ``Hadamard test.'' For $k=(i,j)\in\{1,\ldots,N\}$ such that $i<j$ let $R^\com_k$ be the relation $[x_i,x_j]=e$.
Then we define
\begin{equation}\label{eq:z2k-eff}
 G' \,=\,\big\langle x_1,\ldots,x_N \;:\; \{R^\sq_k\}\,,\; \{R^\ld_k\}\,,\; \{R^\had_{k}\}\, , \; \{ R^\com_k\}\big\rangle\;.
\end{equation}
From Lemma~\ref{lem:com-code} it follows that $G'=\Z_2^K$. Our main result is an efficient stability result for this presentation. To state this we need to introduce distributions $\mu_S$ and $\mu_R$ on the generators and relations of $G'$. The distribution $\mu_S$ is taken to be the uniform distribution over $[N] = (\F_2^t)^m \times \F_2^t$. The distribution $\mu_R$ is obtained as follows. With probability $1/4$ each, a relation from $\{R^\sq_k\}$, $\{R^\ld_k\}$ or $\{R^\had_k\}$ is chosen uniformly at random. With probability $1/4$, a random commutation relation from $R^\com_k$ is chosen according to the uniform mixture of the following two distributions:
\begin{enumerate}
\item For the first distribution, we select $u\in \F_q^m$ uniformly at random, $j\in\{1,\ldots,m\}$ uniformly at random, and $i\neq i'\in\{0,\ldots,d\}$ uniformly at random. Finally, select $\alpha,\beta\in \F_2^t$ uniformly at random and check commutation between $x_{u+i e_j,\alpha}$ and $x_{u+i' e_j,\beta}$. 
\item For the second distribution, we first select $j\in\{1,\ldots,m\}$ and $u_{m-j+2},\ldots,u_m \in \F_2^t$ uniformly at random. Then select $v,v' \in (\F_2^t)^m$ uniformly at random, conditioned on the last $(j-1)$ coordinates matching $u_{m-j+2},\ldots,u_m$. Finally, select $\alpha,\beta\in \F_2^t$ uniformly at random and check commutation between $x_{u,\alpha}$ and $x_{u,\beta}$. 
\end{enumerate}

@@

\begin{proof}
Let $(\mM,\tau)$ be a tracial von Neumann algebra and $\phi$ be an $\eps$-homomorphism of $\langle S:R\rangle$ on $(\mU(\mM),\tau)$. 
 Here, $S = \{s_{u,\alpha}: u\in (\F_2^t)^m, \alpha\in \F_2^t\}$ and $R$ is the set of all relations in~\eqref{eq:z2k-eff}, i.e.\ $R= \{R_k^\sq\}\cup \{R^\ld_k\}\cup\{R^\had_k\}\cup\{R^\com_k\}$. Let $N=|S|$.

@@

\begin{claim}\label{claim:z2-stab-1}
There is an $\eps^{(1)}=O(d\eps)$ and an $\eps^{(1)}$-homomorphism $\phi^{(1)}$ of $\langle S:R\rangle$ on $(\mU(\mM),\tau)$ such that $\phi^{(1)}(x)$ is a Hermitian involution for all $x\in S$, and furthermore
\begin{equation}\label{eq:z2-stab-1a}
 \Es{x\sim\mu_S} \big\| \phi(x) - \phi^{(1)}(x) \big\|_\tau^2 \,\leq\, \eps^{(1)}\;.
\end{equation}
\end{claim}

\begin{proof}
Using elementary calculations (see e.g.~\cite[Lemma 3.6]{slofstra2019set}) we see that for any complex $\alpha$, 
\[ \big| \sgn\Re\alpha-\alpha\big| \,\leq\, \Big(1+\frac{1}{\sqrt{2}}\Big) \big|\alpha^2 -1 \big|\;.\]
For any $i\in \{1,\ldots,N\}$ let $\phi^{(1)}(x_i) = \sgn\Re (\phi(x_i))$. Then the claim follows since 
\[ \Es{i\in\{1,\ldots,N\}} \big\| \phi(x_i)^2-\Id \big\|_\tau^2 \,\leq\, 4\eps\;,\]
by assumption and the definition of $\mu_R$, which places weight $\frac{1}{4}$ on the relations in $R_k^\sq$. Since $\mu_S$ is uniform on $S$,~\eqref{eq:z2-stab-1a} follows. Furthermore, because all relations in $R$ have length at most $O(d)$, $\phi^{(1)}$ is an $O(d\eps)$-homomorphism of $\langle S:R\rangle$\footnote{This requires more explanation, e.g.\ we need the marginal of any variable appearing in a relation to be uniform. I'm also not sure it's not $d^2$ instead of $d$.}. 
\end{proof}

@@

\begin{claim}\label{claim:z2-stab-2}
Let $q=2^t$. 
For every $u\in \F_q^m$ there is a projective measurement $\{P^u_\alpha\}_{\alpha\in \F_{q}}$ on $\mM$ such that 
\begin{equation}\label{eq:z2-stab-2}
 \Es{u\in \F_q^m} \Es{\alpha\in \F_2^t} \Big\| \phi(s_{u,\alpha}) - \sum_{\beta\in\F_q} (-1)^{\alpha \cdot \kappa(\beta)} P^u_\beta \Big\|_\tau^2 \,=\, O(\eps)\;. 
\end{equation}
\end{claim}

\begin{proof}
Since $\mu_R$ places weight $1/4$ on relations $\{R^\had_k\}$ we deduce that 
\begin{equation}\label{eq:stab-rm-1}
\Es{u\in \F_q^m} \Es{\alpha,\beta\in \F_2^t} \big\|\phi(s_{u,\alpha})\phi(s_{u,\beta})\phi(s_{u,\alpha+\beta})-\Id\big\|_\tau^2 \,\leq\, 4\eps\;. 
\end{equation}
Fix an $u\in \F_q^m$ and apply Lemma~\ref{lem:had-stab} for that $u$. This gives a partial isometry $w\in P\mM_\infty\Id_\mM$ and a representation $\{U_\alpha\}$ of $\F_2^t$ on $\mN=P\mM_\infty P$ such that 
\begin{equation}\label{eq:stab-rm-1b}
 \Es{\alpha \in \F_2^t} \big\| \phi(s_{u,\alpha}) - w^* U_\alpha w \big\|_\tau^2 \,=\, O(\eps_u)\;,
\end{equation}
where $\eps_u\geq 0$ is such that $\Es{u\in \F_q^m} \eps_u = 4\eps$. Because $\{U_\alpha\}$ are a representation of the abelian group $\F_2^t$, there is a projective measurement $\{Q^u_\beta\}_{\beta\in \F_2^t}$ on $\mN$ such that $U_\alpha = \sum_\beta (-1)^{\alpha\cdot \beta} Q^u_\beta$. ($\{Q^u_\beta\}$ can be found explicitly by applying the Fourier transform.) Applying Lemma~\ref{lem:pull-back}, we deduce a projective measurement $\{P^u_\beta\}_{\beta\in \F_q}$  on $\mM$ such that (by the triangle inequality)
\begin{align*}
 \Es{\alpha \in \F_2^t} \Big\| \phi(s_{u,\alpha}) - \sum_{\beta\in\F_q} (-1)^{\alpha \cdot \kappa(\beta)} P^u_\beta \Big\|_\tau^2
&\leq 2 \Es{\alpha \in \F_2^t} \Big( \Big\| \phi(s_{u,\alpha}) - w^*\Big(\sum_{\beta\in\F_2^t} (-1)^{\alpha \cdot (\beta)} Q^u_\beta \Big) w \Big\|_\tau^2 \\
&\qquad\qquad+   \Big\|  \sum_{\beta\in\F_2^t} (-1)^{\alpha \cdot \beta} w^* Q^u_\beta w - \sum_{\beta\in\F_q} (-1)^{\alpha \cdot \kappa(\beta)} P^u_\beta \Big\|_\tau^2\Big)\\
 &\leq\, O(\eps_u) +  \sum_{\beta\in\F_q} \Big\| w^* Q^u_{\kappa(\beta)} w - P^u_{\beta} \Big\|_\tau^2\\
&\leq O(\eps_u)\;,
\end{align*}
where the first inequality follows from the triangle inequality, the second is by~\eqref{eq:stab-rm-1b} for the first term and Parseval's identity for the second, and the last by the guarantees obtained from Lemma~\ref{lem:pull-back}.
\end{proof}

>>>>

\begin{lemma}\label{lem:pull-back}
Let  $(\mM,\tau^\mM)$ be a tracial von Neumann algebra, $P\in\mM_\infty$ a projection of finite trace, $\mN=P\mM_\infty P$ and $\tau^\mN=\tau_\infty/\tau_\infty(P)$, and $w\in P \mM_\infty \Id_\mM$ a partial isometry. Let 
\[ \eps = \max\big\{ \tau^\mM\big(\Id_\mM - w^* w\big)\,,\;\tau^\mN\big( P- w w^*\big)\big\}\;.\] 
 Then for any projective measurement $\{T_a\}_{a \in \mA}$ on $\mN$, there is a projective measurement $\{Q_a\}_{a \in \mA}$ on $\mM$ such that 
\begin{equation}
\label{eq:pull-back} \sum_{a \in \mA} \big\| Q_a - w^* T_a w\big\|_\tau^2 \,\leq \ 56\eps\;.
\end{equation}
\end{lemma}

@@

\begin{definition}[Almost homomorphism]
Let $G = \langle S:R\rangle $ be a finitely presented group, $\mu$ a distribution on $R$, and $(\mM,\tau)$ a tracial von Neumann algebra. An $(\eps,\mu)$-almost homomorphism of $G$ on $(\mM,\tau)$ is a homomorphism $\phi:\mF(S)\to\mU(\mM)$ such that
\[ \Es{r\sim \mu} \big\|  \phi(r) - \Id \big \|_\tau^2 \,\leq\, \eps\;.\]
\end{definition}

@@

Let $G_\bRM = G(h_\bRM)$ be the group that is presented from $h_\bRM$ (\hnote{added:} recall from \Cref{sec:pres-code} that codes give rise to group presentations). Explicitly, $h_\bRM$ is the parity check matrix that arises from the tester described in Figure~\ref{fig:bRM-tester}.

@@

We do not know if $G_\bRM = \Z_2^K$, with $K=t(d+1)^m$. Instead we modify the presentation $G(h_\bRM)$ by adding pairwise commutation relations in a similar manner as Lemma~\ref{lem:com-code}. Let 
\[ G(h_\bRM) \,=\, \big\langle x_1,\ldots,x_N \;:\; \{R^\sq_k\}\,,\; \{R^\ld_k\}\,,\; \{R^\had_{k}\} \big\rangle\;,\]
where  $R^\sq_k$ ranges overal all relations of the form $x_i^2=e$ for $i\in \{1,\ldots,N\}$, $R^\ld_k$ ranges over all relations implied by the ``low-degree test'' in Figure~\ref{fig:bRM-tester} and $R^\had_{k}$ ranges over all the relations implied by the ``Hadamard test.'' For $k=(i,j)\in\{1,\ldots,N\}$ such that $i<j$ let $R^\com_k$ be the relation $[x_i,x_j]=e$.
Then we define
\begin{equation}\label{eq:z2k-eff}
 G' \,=\,\big\langle x_1,\ldots,x_N \;:\; \{R^\sq_k\}\,,\; \{R^\ld_k\}\,,\; \{R^\had_{k}\}\, , \; \{ R^\com_k\}\big\rangle\;.
\end{equation}
From Lemma~\ref{lem:com-code} it follows that $G'=\Z_2^K$. Our main result is an efficient stability result for this presentation. To state this we need to introduce distributions $\mu_S$ and $\mu_R$ on the generators and relations of $G'$. The distribution $\mu_S$ is taken to be the uniform distribution over $[N] = (\F_2^t)^m \times \F_2^t$. The distribution $\mu_R$ is obtained as follows. With probability $1/4$ each, a relation from $\{R^\sq_k\}$, $\{R^\ld_k\}$ or $\{R^\had_k\}$ is chosen uniformly at random. With probability $1/4$, a random commutation relation from $R^\com_k$ is chosen according to the uniform mixture of the following two distributions:
\begin{enumerate}
\item For the first distribution, we select $u\in \F_q^m$ uniformly at random, $j\in\{1,\ldots,m\}$ uniformly at random, and $i\neq i'\in\{0,\ldots,d\}$ uniformly at random. Finally, select $\alpha,\beta\in \F_2^t$ uniformly at random and check commutation between $x_{u+i e_j,\alpha}$ and $x_{u+i' e_j,\beta}$. 
\item For the second distribution, we first select $j\in\{1,\ldots,m\}$ and $u_{m-j+2},\ldots,u_m \in \F_2^t$ uniformly at random. Then select $v,v' \in (\F_2^t)^m$ uniformly at random, conditioned on the last $(j-1)$ coordinates matching $u_{m-j+2},\ldots,u_m$. Finally, select $\alpha,\beta\in \F_2^t$ uniformly at random and check commutation between $x_{u,\alpha}$ and $x_{u,\beta}$.  \hnote{$v$ and $v'$ don't seem to show up, is it supposed to be $x_{v,\alpha}$ and $x_{v',\beta}$?}
\end{enumerate}

@@

\begin{proof}
Let $(\mM,\tau)$ be a tracial von Neumann algebra and $\phi$ be an $\eps$-homomorphism of $\langle S:R\rangle$ on $(\mU(\mM),\tau)$ \hnote{this should be ``...on $(\mM,\tau)$'', right? (see def of homomorphism)}. 
 Here, $S = \{s_{u,\alpha}: u\in (\F_2^t)^m, \alpha\in \F_2^t\}$ \hnote{what is $s_{u,\alpha}$? should it be $x_{u,\alpha}$?} and $R$ is the set of all relations in~\eqref{eq:z2k-eff}, i.e.\ $R= \{R_k^\sq\}\cup \{R^\ld_k\}\cup\{R^\had_k\}\cup\{R^\com_k\}$. Let $N=|S|$.

@@

\begin{claim}\label{claim:z2-stab-1}
There is an $\eps^{(1)}=O(d\eps)$ and an $\eps^{(1)}$-homomorphism $\phi^{(1)}$ of $\langle S:R\rangle$ on $(\mU(\mM),\tau)$ \hnote{same thing, should be $(\mM,\tau)$ I think} such that $\phi^{(1)}(x)$ is a Hermitian involution for all $x\in S$, and furthermore
\begin{equation}\label{eq:z2-stab-1a}
 \Es{x\sim\mu_S} \big\| \phi(x) - \phi^{(1)}(x) \big\|_\tau^2 \,\leq\, \eps^{(1)}\;.
\end{equation}
\end{claim}

\begin{proof}
Using elementary calculations (see e.g.~\cite[Lemma 3.6]{slofstra2019set}) we see that for any complex $\alpha$, 
\[ \big| \sgn\Re\alpha-\alpha\big| \,\leq\, \Big(1+\frac{1}{\sqrt{2}}\Big) \big|\alpha^2 -1 \big|\;.\]
For any $i\in \{1,\ldots,N\}$ let $\phi^{(1)}(x_i) = \sgn\Re (\phi(x_i))$ \hnote{maybe there should be a comment somewhere saying we identify $\{1,\ldots,N\}$ with $(\F_2^t)^m \times \F_2^t$}. Then the claim follows since 
\[ \Es{i\in\{1,\ldots,N\}} \big\| \phi(x_i)^2-\Id \big\|_\tau^2 \,\leq\, 4\eps\;,\]
by assumption and the definition of $\mu_R$, which places weight $\frac{1}{4}$ on the relations in $R_k^\sq$. Since $\mu_S$ is uniform on $S$,~\eqref{eq:z2-stab-1a} follows. Furthermore, because all relations in $R$ have length at most $O(d)$, $\phi^{(1)}$ is an $O(d\eps)$-homomorphism of $\langle S:R\rangle$\tnote{This requires more explanation, e.g.\ we need the marginal of any variable appearing in a relation to be uniform. I'm also not sure it's not $d^2$ instead of $d$.} \hnote{What is the length of a relation? The number of variables in it?}. 
\end{proof}

@@

\begin{claim}\label{claim:z2-stab-2}
Let $q=2^t$. 
For every $u\in \F_q^m$ there is a projective measurement $\{P^u_\alpha\}_{\alpha\in \F_{q}}$ on $\mM$ such that 
\begin{equation}\label{eq:z2-stab-2}
 \Es{u\in \F_q^m} \Es{\alpha\in \F_2^t} \Big\| \phi(s_{u,\alpha}) - \sum_{\beta\in\F_q} (-1)^{\alpha \cdot \kappa(\beta)} P^u_\beta \Big\|_\tau^2 \,=\, O(\eps)\;. 
\end{equation}
\hnote{notational consistency: are we using $s_{u,\alpha}$ or $x_{u,\alpha}$?}
\end{claim}

\begin{proof}
Since $\mu_R$ places weight $1/4$ on relations $\{R^\had_k\}$ we deduce that 
\begin{equation}\label{eq:stab-rm-1}
\Es{u\in \F_q^m} \Es{\alpha,\beta\in \F_2^t} \big\|\phi(s_{u,\alpha})\phi(s_{u,\beta})\phi(s_{u,\alpha+\beta})-\Id\big\|_\tau^2 \,\leq\, 4\eps\;. 
\end{equation}
Fix an $u\in \F_q^m$ and apply Lemma~\ref{lem:had-stab} for that $u$. This gives a partial isometry $w\in P\mM_\infty\Id_\mM$ and a \hnote{added:} unitary representation $\{U_\alpha\}$ of $\F_2^t$ on $\mN=P\mM_\infty P$ such that 
\begin{equation}\label{eq:stab-rm-1b}
 \Es{\alpha \in \F_2^t} \big\| \phi(s_{u,\alpha}) - w^* U_\alpha w \big\|_\tau^2 \,=\, O(\eps_u)\;,
\end{equation}
where $\eps_u\geq 0$ is such that $\Es{u\in \F_q^m} \eps_u = 4\eps$. Because $\{U_\alpha\}$ are a representation of the abelian group $\F_2^t$, there is a projective measurement $\{Q^u_\beta\}_{\beta\in \F_2^t}$ on $\mN$ such that $U_\alpha = \sum_\beta (-1)^{\alpha\cdot \beta} Q^u_\beta$. ($\{Q^u_\beta\}$ can be found explicitly by applying the Fourier transform. \hnote{should we just directly say that $Q^u_\beta = \sum_\alpha (-1)^{\alpha \cdot \beta} U_\alpha$ is a projective measurement?}) Applying Lemma~\ref{lem:pull-back}, we deduce a projective measurement $\{P^u_\beta\}_{\beta\in \F_q}$  \hnote{here (and in what follows) it's kind of confusing because $\beta$ keeps switching back and forth between $\F_q$ and $\F_q^t$, which means that sometimes it's $\kappa(\beta)$ and $\beta$, which is confusing...} on $\mM$ such that (by the triangle inequality)
\begin{align*}
 \Es{\alpha \in \F_2^t} \Big\| \phi(s_{u,\alpha}) - \sum_{\beta\in\F_q} (-1)^{\alpha \cdot \kappa(\beta)} P^u_\beta \Big\|_\tau^2
&\leq 2 \Es{\alpha \in \F_2^t} \Big( \Big\| \phi(s_{u,\alpha}) - w^*\Big(\sum_{\beta\in\F_2^t} (-1)^{\alpha \cdot (\beta)} Q^u_\beta \Big) w \Big\|_\tau^2 \\
&\qquad\qquad+   \Big\|  \sum_{\beta\in\F_2^t} (-1)^{\alpha \cdot \beta} w^* Q^u_\beta w - \sum_{\beta\in\F_q} (-1)^{\alpha \cdot \kappa(\beta)} P^u_\beta \Big\|_\tau^2\Big)\\
 &\leq\, O(\eps_u) +  \sum_{\beta\in\F_q} \Big\| w^* Q^u_{\kappa(\beta)} w - P^u_{\beta} \Big\|_\tau^2\\
&\leq O(\eps_u)\;,
\end{align*}
where the first inequality follows from the triangle inequality, the second is by~\eqref{eq:stab-rm-1b} for the first term and Parseval's identity for the second, and the last by the guarantees obtained from Lemma~\ref{lem:pull-back}.
\end{proof}
